---
title: "lab10: Halloween Mini Project"
author: 'Ethan Ashley (PID: A15939817)'
date: "2025-10-31"
output: 
  pdf_document
toc: true
---

# Importing the 538 Candy Popularity Dataset

```{r}
candy_file <- "candy-data.csv"

candy = read.csv(candy_file, row.names = 1)

#head(candy)
```


**Q1. How many different candy types are in this dataset?**

There are 85 different types of candy.

```{r}
nrow(candy)
```

**Q2. How many fruity candy types are in the dataset?**

There are 38 fruity candy types in the dataset.

```{r}
sum(candy$fruity)
```

# What is your favorite candy?

**Q3. What is your favorite candy in the dataset and what is it’s winpercent value?**

My favorite candy in the dataset is M&M's. The win percent for M&M's is 66.6%.
  
```{r}
candy["M&M's", ]$winpercent
```

**Q4. What is the winpercent value for “Kit Kat”?**

The winpercent value for Kit Kat is 76.8%.

```{r}
candy["Kit Kat", ]$winpercent
```

**Q5. What is the winpercent value for “Tootsie Roll Snack Bars”?**

The winpercent value for Tootsie Roll Snack Bars is 49.7.

```{r}
candy["Tootsie Roll Snack Bars", ]$winpercent
```
*`skim(candy)` could not be included because of a unicode support error with installation of TinyTex/Latex*
```{r}
library("skimr")
#skim(candy)
```


**Q6. Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?**

The winpercent column seems to be on a different scale than the other columns in the dataset.

**Q7. What do you think a zero and one represent for the candy$chocolate column?**

I think zero and one represent whether a candy has chocolate or not.

**Q8. Plot a histogram of winpercent values**

```{r}
library(ggplot2)

ggplot(candy) + aes(winpercent) + geom_histogram(bins = 15)
```

**Q9. Is the distribution of winpercent values symmetrical?**

The distribution of winpercent values is not symmetrical.

```{r}
ggplot(candy) + aes(winpercent) + geom_density()
```

**Q10. Is the center of the distribution above or below 50%?**

The center of the distribution appears to be below 50%.

```{r}
summary(candy$winpercent)
```

**Q11. On average is chocolate candy higher or lower ranked than fruit candy?**

On average chocolate candy is higher ranked than fruity candy.

```{r}
chocWinPercent <- candy$winpercent[as.logical(candy$chocolate)]
fruityWinPercent <- candy$winpercent[as.logical(candy$fruity)]

mean_chocWinPerecent <- mean(chocWinPercent)
mean_fruityWinPercent <- mean(fruityWinPercent)

mean_chocWinPerecent > mean_fruityWinPercent
```

**Q12. Is this difference statistically significant?**

Using a T-test, the difference is statistically significant.

```{r}
t.test(chocWinPercent, fruityWinPercent)
```

# Overall Candy Rankings

**Q13. What are the five least liked candy types in this set?**

The five least liked candies in this set are Nik L Nip, Boston Baked Beans, Chiclets, Super Bubble, and Jawbusters.

```{r}
library(dplyr)

candy %>% arrange(winpercent) %>% head(5)
```

**Q14. What are the top 5 all time favorite candy types out of this set?**

The top 5 most popular candies are in this set are Reese's Peanut Butter Cups, Reese's Miniatures, Twix, Kit Kat, and Snickers.

```{r}
candy %>% arrange(desc(winpercent)) %>% head(5)
```

**Q15. Make a first barplot of candy ranking based on winpercent values.**

```{r}
ggplot(candy) + aes(winpercent, rownames(candy)) + geom_col() + theme(axis.text=element_text(size=4))
```

**Q16. This is quite ugly, use the reorder() function to get the bars sorted by winpercent?**

```{r}
ggplot(candy) + aes(winpercent, reorder(rownames(candy),winpercent)) + geom_col() + theme(axis.text=element_text(size=4))
```

```{r}
my_cols=rep("black", nrow(candy))
my_cols[as.logical(candy$chocolate)] = "chocolate"
my_cols[as.logical(candy$bar)] = "brown"
my_cols[as.logical(candy$fruity)] = "pink"

ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy),winpercent)) +
  geom_col(fill=my_cols) + theme(axis.text=element_text(size=4))
```

**Q17. What is the worst ranked chocolate candy?**

The worst ranked chocolate candy is Sixlets.

**Q18. What is the best ranked fruity candy?**

The best ranked fruit candy is Starbursts.

# Taking a look at pricepercent

```{r}
library(ggrepel)

# How about a plot of price vs win
ggplot(candy) +
  aes(winpercent, pricepercent, label=rownames(candy)) +
  geom_point(col=my_cols) + 
  geom_text_repel(col=my_cols, size=3.3, max.overlaps = 5)
```

**Q19. Which candy type is the highest ranked in terms of winpercent for the least money - i.e. offers the most bang for your buck?**

The candy type that is highest ranked for the best price is Reese's Miniatures.

**Q20. What are the top 5 most expensive candy types in the dataset and of these which is the least popular?**

The top 5 most expensive candy types are Nik L Nip, Nestle Smarties, Ring pop, Hershey's Krackel, and Hershey's Milk Chocolate.

The least popular of these candies is Nik L Nip.

```{r}
candy %>% arrange(desc(candy$pricepercent)) %>% head(5)
```

# Exploring the correlation structure

```{r}
library(corrplot)

cij <- cor(candy)
corrplot(cij)
```

**Q22. Examining this plot what two variables are anti-correlated (i.e. have minus values)?**

The chocolate and fruity variables are anti-correlated.

**Q23. Similarly, what two variables are most positively correlated?**

The two most positively correlated variables are chocolate and winpercent.

# Principal Component Analysis

```{r}
pca <- prcomp(candy, scale = TRUE)
summary(pca)
```

```{r}
plot(pca$x[,1:2])
```

```{r}
plot(pca$x[,1:2], col=my_cols, pch=16)
```

```{r}
# Make a new data-frame with our PCA results and candy data
my_data <- cbind(candy, pca$x[,1:3])

p <- ggplot(my_data) + 
        aes(x=PC1, y=PC2, 
            size=winpercent/100,  
            text=rownames(my_data),
            label=rownames(my_data)) +
        geom_point(col=my_cols)

p
```

```{r}
p + geom_text_repel(size=3.3, col=my_cols, max.overlaps = 7)  + theme(legend.position = "none") + labs(title="Halloween Candy PCA Space", subtitle="Colored by type: chocolate bar (dark brown), chocolate other (light brown), fruity (red), other (black)", caption="Data from 538")
```

**plotly element was not run in the final report because it could not be rendered into pdf**

```{r}
library(plotly) #useful for producing interactive plots that can be integrated in html outputs

#ggplotly(p)
```


```{r}
par(mar=c(8,4,2,2))
barplot(pca$rotation[,1], las=2, ylab="PC1 Contribution")
```

**Q24. What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you?**

The variables fruity, hard, and, pluribus are picked up strongly in PC1. These make sense to me, most people don't prefer harder, fruit flavored candies.
