---
title: 'Class06: Functions Lab'
author: 'Ethan Ashley (PID: A15939817)'
date: "2025-10-17"
output:
  pdf_document: default
#toc: TRUE
---

# Introduction

All functions in R have at least 3 things"

- A **name**, we pick this and use it to call our function
- Input **arguments** (there can be multiple)
- The **body** lines of R code that do the work

## Our first function

```{r}
add <- function(x, y=1) {
  x + y
}
```

Now call this function:
```{r}
add(10)
```


## A second function

Write a function to generate random nucleotide sequences of a user specified length:

The `sample()` function can be helpful here. It randomly samples from a supplied vector.

```{r}
randSeq <- function(len) {
  bases = c("A", "C", "G", "T")
  paste(sample(bases, len, replace = TRUE), collapse="")
  #paste can concatenate vectors of characters into a single output
}
```

```{r}
randSeq(50)
```


## Flow control

```{r}
fasta <- FALSE
if(fasta) {
  cat("Hello you!")
} else {
  cat("No you dont!")
}
```

Add the ability to return a multi-element vector or a single element fasta like vector from the random DNA sequence generating function.

```{r}
randSeqFormat <- function(len, fasta=TRUE) {
  #if format is set to true, then the vector will be returned in a fasta format
  bases <- c("A", "C", "G", "T")
  seq <- sample(bases, len, replace = TRUE)
  
  if(fasta) {
    #paste can concatenate vectors of characters into a single output
    paste(seq, collapse="")
  } else {
    seq
  }
}
```

```{r}
seq <- randSeqFormat(48, TRUE)
seq
```
## A Protein Generating Function

```{r}
generateProtein <- function(len, fasta=TRUE){
  amino_acids = c("A", "R", "N", "D", "E", "G", "H", "I", "L", "K", "F", "P", "S", "T", "W", "V", "C", "Y")
  
  seq <- sample(amino_acids, len, replace = TRUE)
  
  if(fasta) {
    #paste can concatenate vectors of characters into a single output
    return(paste(seq, collapse=""))
  } else {
    return(seq)
  }
}
```


Use this new function to make random protein sequences of length 6 to 12 (i.e., one w/ 6 AA's, one w/ 7 AA's, etc.)

```{r}
#lengths <- 6:12

for(i in 6:12) {
  cat(">", i, "\n", sep="")
  cat(generateProtein(i))
  cat("\n")
}
```


A better way to solve this is to use the `apply()` family of functions, specifically the `sapply()` function.

```{r}

sapply(6:12, generateProtein)

```