---
title: "HW06"
author: 'Ethan Ashley (PID: A15939817)'
date: "2025-10-17"
output: pdf_document
toc: true
---

# Generalize the following protein analysis code to work for any protein sequence

```{r}
# Can you improve this analysis code?
library(bio3d)
s1 <- read.pdb("4AKE") # kinase with drug
s2 <- read.pdb("1AKE") # kinase no drug
s3 <- read.pdb("1E4Y") # kinase with drug

s1.chainA <- trim.pdb(s1, chain="A", elety="CA")
s2.chainA <- trim.pdb(s2, chain="A", elety="CA")
s3.chainA <- trim.pdb(s1, chain="A", elety="CA")

s1.b <- s1.chainA$atom$b
s2.b <- s2.chainA$atom$b
s3.b <- s3.chainA$atom$b

plotb3(s1.b, sse=s1.chainA, typ="l", ylab="Bfactor")
plotb3(s2.b, sse=s2.chainA, typ="l", ylab="Bfactor")
plotb3(s3.b, sse=s3.chainA, typ="l", ylab="Bfactor")
```


## Q6. How would you generalize the original code above to work with any set of input protein structures?

Converting the general code above into a more generalized version

```{r}
inputProteins = c("4AKE", "1AKE", "1E4Y")

#generating a for loop that takes a list of PDB protein ID's and analyzes each one individually as shown above

for (i in inputProteins) {
  s <- read.pdb(i)
  s.chainA <- trim.pdb(s, chain="A", elety="CA")
  s.b <- s.chainA$atom$b
  plotb3(s.b, sse=s.chainA, typ="l", ylab="Bfactor")
}
```


Converting the generalized code into a function that can be applied to any list of sequences

```{r}

#This function accepts a list of PDB protein ID's and analyzes each one individually as shown above. main is added as an input to plotb3 to clarify which protein is plotted.

analyzeProteins <- function(seqs) {
  for (i in inputProteins) {
    s <- read.pdb(i)
    s.chainA <- trim.pdb(s, chain="A", elety="CA")
    s.b <- s.chainA$atom$b
        plotb3(s.b, sse=s.chainA, typ="l", ylab="Bfactor", main=i)
  }
}
```

**This function accepts a list of PDB protein ID's and analyzes each one individually as shown above generating a plot of the residue position vs. the B factor**


Testing new function on the provided list of sequences
```{r}
analyzeProteins(inputProteins)
```

## Testing with `sapply()`

**Making the function even more generalized so that it can be utilized with sapply(). It accepts a single PDB ID at a time but can be vectorized with sapply() to achieve the same output as previously**

```{r}
analyzeProteins <- function(seqs) {
    s <- read.pdb(i)
    s.chainA <- trim.pdb(s, chain="A", elety="CA")
    s.b <- s.chainA$atom$b
    plotb3(s.b, sse=s.chainA, typ="l", ylab="Bfactor", , main=i)
}
```

```{r}
sapply(inputProteins, analyzeProteins)
```

